PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-04-18 papers/NAR_web_server_2009/paper.tex
Modifications to introduction of response to reviewers

% Some shortcuts to make life easier when writing latex
\newcommand{\peace} {{\small PEACE}}
\newcommand{\wcd} {{\small WCD}}
\newcommand{\capthree} {{\small Cap3}}

@@

\begin{figure}
  \centerline{\includegraphics[width=3in]{screen.d/workflow.pdf}}
  \caption{Overview of the procedure for clustering and analysis using
    PEACE.}\label{fig:workflow}
\end{figure}

@@

\begin{figure*}
  \begin{minipage}{3in}
    \includegraphics[width=3in]{screen.d/install_page_big.png}
    \centerline{\small{(a)}}
  \end{minipage}
  \begin{minipage}{2in}
    \vspace{0.47in}
    \includegraphics[width=2in]{screen.d/add_server_big.png}
    \centerline{\small{(b)}}
  \end{minipage}
  \begin{minipage}{2in}
    \vspace{0.66in}
    \includegraphics[width=2in]{screen.d/processor_assignment_big.png}
    \centerline{\small{(c)}}
  \end{minipage}

  \begin{minipage}{3in}
    \includegraphics[width=3in]{screen.d/cluster_list_big.png}
    \centerline{\small{(d)}}
  \end{minipage}
  \begin{minipage}{1in}
    \hspace{1in}
  \end{minipage}
  \begin{minipage}{3in}
    \includegraphics[width=3in]{screen.d/classifier_big.png}
    \centerline{\small{(e)}}
  \end{minipage}

@@

\peace\/ has been tested on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
\capthree\/ assembly tool \cite{Huang99} (the latter of which
implicitly calculates a clustering in the process of assembly).  For our
simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03} to
generate simulated transcript fragments of varying length under
different models of error (Supplementary Materials, Section~C.1),
generating the fragments from the list of 100 zebra fish genes used in
the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
to match, as closely as possible, those used in the \wcd\/ study (see
Supplementary Materials).  Our primary methods of quality assessment
were {\it sensitivity} (the fraction of fragment pairs from the same
gene that were correctly clustered together) and {\it Type 1 error}
(the fraction of genes that were divided between clusters)
\cite{Wang04,Hazelhurst08a}.  Specificity was not an issue in
simulated tests, as both \peace\/ and \wcd\/ were completely successful
in separating unrelated ESTs (save when dealing with recently
duplicated genes).  Other measurements are also discussed in the
supplementary materials.  In Figure~\ref{SeT1} (left), reflecting
tests done on simulated Sanger EST sets, we see a
significant improvement of \peace\/ (blue) over \wcd\/ (green) and
\capthree\/ (black) in its sensitivity to sequencing errors, while in
Figure~\ref{SeT1} (right) we see a comparable improvement in Type 1
error.  In Figure~\ref{singletons} we look at the number of singleton
clusters (fragments not joined to any cluster), which should not occur
in our simulated sets.  We again see significant improvement in \peace.
The sequential runtime of \peace\/ is slightly improved over that of
\wcd\/ (see Supplementary Materials, Figure~S4).

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeT1.pdf}}
  \caption{Comparisons of Sensitivity and Type 1 error, based on the
    average over 30 simulated Sanger Sequence ESTs sets derived from 100 zebra
    fish genes  (see
    Supplementary Materials, Section C, for more details).  Blue/Solid
    = \peace, Green/Dash = \wcd, Black/Dot-Dash = \capthree; vertical
    tics = 95\% confidence intervals on estimates.  Intervals are not
    presented for Type 1 error due to the extremely small
    variance.}\label{SeT1}
\end{figure}

\begin{figure}
  \centerline{\includegraphics[width=3in]{pics.d/singletons.pdf}}
  \caption{Average number of fragments flagged as singletons by each tool
    when run on the simulated sequences; correct answer in all cases is
    zero.  Blue/Solid = \peace, Green/Dash = \wcd,
    Black/Dot-Dash = \capthree.}\label{singletons}
\end{figure}

In applying the tools to real data, we started with a set of
approximately $190,000$ clean Sanger ESTs derived from the {\it
  Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
fragment set to the genome, taking this as our reference clustering.
We see slight improvements in \peace\/ over \wcd\/ in both sensitivity
and Type 1 error (both significantly outperforming \capthree).  Using
the mouse data set used in Hazelhurst {\it et al.}
\cite{Hazelhurst08a} \peace\/ again shows slight improvements in
sensitivity and Type 1 error rates, with an 18\% speedup for \peace\/ 
(see Supplementary Materials, Section~D, for more details).

We were unable to run \wcd\/ on short-read sequences, as the
distributed implementation is unable to handle such data without an
adjustment to the code.  Hence, while \peace\/ was run on both
simulated and real NGS fragment sets, there was no cluster tool to
which the results could be compared.  \peace\/ did well, achieving a
sensitivity of 0.75 when applied to a simulated short-read set subjected to
a sequencing error rate of 2\% (with comparable {\it Jaccard Index}
values reflecting both sensitivity {\it and} specificity
\cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was approximately 17.6
minutes to process the 7.9 Mb of data when run in sequential mode on a
3.2 GHz Intel Xeon CPU.  When applied to a 64 Mb set of 165,490 {\it
  Chlamydomonas reinhardtii} transcript fragments sequenced by 454
pyrosequencing technology (ranging in size from 50 bases to several
hundred), \peace\/ completed the job in 49 minutes using 30 of the
Xeon processors.

>>>>

% Some shortcuts to make life easier when writing latex
\newcommand{\peace} {{\small PEACE}}
\newcommand{\wcd} {{\small WCD}}
\newcommand{\capthree} {{\small Cap3}}
\newcommand{\easycluster} {{\small EasyCluster}}

@@

\begin{figure}
  \centerline{\includegraphics[width=3in]{pics.d/workflow.eps}}
  \caption{Overview of the procedure for clustering and analysis using
    PEACE.}\label{fig:workflow}
\end{figure}

@@

\begin{figure*}
  \begin{minipage}{3in}
    \includegraphics[width=3in]{install_page_big.eps}
    \centerline{\small{(a)}}
  \end{minipage}
  \begin{minipage}{2in}
    \vspace{0.47in}
    \includegraphics[width=2in]{add_server_big.eps}
    \centerline{\small{(b)}}
  \end{minipage}
  \begin{minipage}{2in}
    \vspace{0.47in}
    \includegraphics[width=2in]{processor_assignment_big.eps}
    \centerline{\small{(c)}}
  \end{minipage}

  \begin{minipage}{3in}
    \includegraphics[width=3in]{cluster_list_big.eps}
    \centerline{\small{(d)}}
  \end{minipage}
  \begin{minipage}{1in}
    \hspace{1in}
  \end{minipage}
  \begin{minipage}{3in}
    \includegraphics[width=3in]{classifier_big.eps}
    \centerline{\small{(e)}}
  \end{minipage}

@@

\peace\/ has been tested on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the \wcd\/ clustering tool \mc{(version
  0.5.1)}\cite{Hazelhurst08a} and the \capthree\/ assembly tool
\cite{Huang99} (the latter of which implicitly calculates a clustering
in the process of assembly).  For our simulation tests we used the
{\bf ESTSim} tool \cite{Hazelhurst03} to generate simulated transcript
fragments of varying length under different models of error
(Supplementary Materials, Section~C.1), generating the fragments from
the list of 100 zebra fish genes used in the \wcd\/ testing
\cite{Hazelhurst08a}.  Tool parameters were taken to match, as closely
as possible, those used in the \wcd\/ study (see Supplementary
Materials).  \mc{The most important method of quality assessment is
  {\it sensitivity} (the fraction of fragment pairs from the same gene
  that were correctly clustered together).  But we also look at the
  {\it Jaccard Index} (which balances sensitivity with the number of
  false positives by looking at the ratio of true positives to the sum
  of true positives, false positives, and true negitives), {\it Type 1
    error} (the fraction of genes that were divided between clusters),
  and {\it Type 2 error} (the fraction of clusters containing two or
  more genes).  Specificity was not an issue in simulated tests, as
  both \peace\/ and \wcd\/ were completely successful in separating
  unrelated ESTs (save when dealing with recently duplicated genes).
  Other measurements are also discussed in the supplementary
  materials.  In Figure~\ref{SeJiT1T2} (left) we see these four tests,...}

% , reflecting tests done on
%   simulated Sanger EST sets, we see a significant improvement of
%   \peace\/ (blue) over \wcd\/ (green) and \capthree\/ (black) in its
%   sensitivity to sequencing errors, while in Figure~\ref{SeT1} (right)
%   we see a comparable improvement in Type 1 error.  In
%   Figure~\ref{singletons} we look at the number of singleton clusters
%   (fragments not joined to any cluster), which should not occur in our
%   simulated sets.  We again see significant improvement in \peace.
%   The sequential runtime of \peace\/ is slightly improved over that of
%   \wcd\/ (see Supplementary Materials, Figure~S4).

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2.40.eps}}
  \caption{Comparisons of Sensitivity and Type 1 error, based on the
    average over 30 simulated Sanger Sequence ESTs sets derived from 100 zebra
    fish genes  (see
    Supplementary Materials, Section C, for more details).  Blue/Solid
    = \peace, Green/Dash = \wcd, Black/Dot-Dash = \capthree; vertical
    tics = 95\% confidence intervals on estimates.  Intervals are not
    presented for Type 1 error due to the extremely small
    variance.}\label{SeJiT1T2}
\end{figure}

\mc{In applying the tools to real data Sanger data, we used the Human
Benchmark Dataset used to test \easycluster\/ \cite{Picardi09}, and
the A076941.fa {\it Arabidopsis thaliana} dataset used to test WCD
\cite{Hazelhurst08,Hazelhurst08a}.  For the first, we find an almost
exact match in results in quality between PEACE and WCD (both
out-preforming Cap3), but a significant gain in runtime for PEACE over
WCD, as shown in Tabe\ref{XXX}}.  For the A076941.fa...}

% we started with a set of
% approximately $190,000$ clean Sanger ESTs derived from the {\it
%   Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
% sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
% fragment set to the genome, taking this as our reference clustering.
% We see slight improvements in \peace\/ over \wcd\/ in both sensitivity
% and Type 1 error (both significantly outperforming \capthree).  Using
% the \mc{Arabidopsis} data set used in Hazelhurst {\it et al.}
% \cite{Hazelhurst08a} \peace\/ again shows slight improvements in
% sensitivity and Type 1 error rates, with an 18\% speedup for \peace\/ 
% (see Supplementary Materials, Section~D, for more details).

We were unable to run \wcd\/ on short-read sequences, as the
distributed implementation is unable to handle such data without an
adjustment to the code.  Hence, while \peace\/ was run on both
simulated and real NGS fragment sets, there was no cluster tool to
which the results could be compared.  \peace\/ did well, achieving a
sensitivity of 0.75 when applied to a simulated short-read set
subjected to a sequencing error rate of 2\% (with comparable {\it
  Jaccard Index} values reflecting both sensitivity {\it and}
specificity \cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was
approximately 17.6 minutes to process the 7.9 Mb of data when run in
sequential mode on a \mc{3.2 GHz Intel Xeon EM64T CPU with a 2 MB
  cache and 800 MHz font side bus}.  When applied to a 64 Mb set of
165,490 {\it
 Chlamydomonas reinhardtii} transcript fragments
sequenced by 454 pyrosequencing technology (ranging in size from 50
bases to several hundred), \peace\/ completed the job in 49 minutes
using 30 of the Xeon processors.
